---
title: "Objective 1.1: Computing biodiversity metrics"
output: html_notebook
---

*Aim: To compute species richness per site, as the sum of species present (from the qualitative, presence/absence site x species matrix).*

Data required: site_species_qualitative_data_matrix.csv

1. Load data and compute richness:
```{r}
library(dplyr)
library(tibble)

qual_data <- read.csv("site_species_qualitative_data_matrix.csv")

comm_pa <- qual_data %>%
  select(-Sr..No., -Flowering.species) %>%
  column_to_rownames("Species.code") %>%
  t()

# Species richness per site: sum of species present (column-wise sums, after transposing)
alpha_richness <- rowSums(comm_pa)

richness_df <- data.frame(
  Site = names(alpha_richness),
  Richness = alpha_richness
)

print(richness_df)
```

*##############################################################################*

**Aim 2: To compute the Shannon-Wiener diversity index (H′) per site, using the quantitative site x species abundance matrix.**

Eriocaulon L. is excluded here, as its consistently high abundance skews indices computed from relative species abundance. Please refer to Supplementary Note S1 for detailed justification.

Data required: site_species_quantitative_data_matrix.csv

2. Load data:
```{r}
library(vegan)

quant_data <- read.csv("site_species_quantitative_data_matrix.csv")
```

3. Prepare the Eriocaulon-excluded matrix and compute Shannon-Wiener diversity:
```{r}
species_matrix_noEriocaulon <- quant_data %>%
  filter(!grepl("Eriocaulon", Flowering.species)) %>%
  select(-Sr..No., -Flowering.species) %>%
  column_to_rownames("Species.code") %>%
  t()

alpha_shannon <- diversity(species_matrix_noEriocaulon, index = "shannon")

shannon_df <- data.frame(
  Site = names(alpha_shannon),
  Shannon_diversity = alpha_shannon
)

print(shannon_df)
```

4. Combine both metrics into a single summary table:
```{r}
alpha_df <- left_join(richness_df, shannon_df, by = "Site")

print(alpha_df)

# write.csv(alpha_df, "species_diversity_per_site.csv", row.names = FALSE)
```

*##############################################################################*
