---
title: "Objective 2.1: Beta diversity partitioning (turnover and nestedness)"
output: html_notebook
---

*Aim: To partition Sørensen's beta diversity (βsor) into its turnover (βsim) and nestedness (βsne) components*

This is to test whether compositional differences among plateaus arise primarily from species replacement or from differential
species loss.

Data required: site_species_qualitative.csv (presence/absence)

1. Load data:
```{r}
library(betapart)
library(vegan)

raw_data <- read.csv("site_species_qualitative_data_matrix.csv")

# Use Species code as row names, drop the full species name column,
# keep only the site presence/absence columns
comm_pa <- raw_data %>%
  select(-Sr..No., -Flowering.species) %>%
  tibble::column_to_rownames("Species.code")

# Transpose: beta.multi() expects sites as rows, species as columns
comm_pa <- t(comm_pa)
```

2. Partition multi-site beta diversity (Sørensen index):
```{r}
beta_parts <- beta.multi(comm_pa, index.family = "sorensen")
beta_parts
```

Result:

beta.SIM → Turnover    : 0.565 : Dissimilarity due to species replacement between sites
beta.SNE → Nestedness  : 0.083 : Dissimilarity due to species loss/gain (nested subsets of richer communities)
beta.SOR → Total       : 0.648 : Total community dissimilarity (replacement + nestedness) across all sites

Interpretation:

1. Total beta diversity (0.648) high compositional dissimilarity across sites.
2. Turnover (0.565) accounts for the majority (87.19%) of this dissimilarity, indicating that most site differences are due to species replacement, not just richness differences.
3. Nestedness (0.083) is relatively small (12.81%), suggesting that species loss or nested subset patterns contribute relatively little to overall beta diversity.

These results suggest that sites host distinct species assemblages, rather than simply being richer or poorer subsets of a regional pool.

3. Plot: relative contribution of turnover and nestedness:
```{r}
library(ggplot2)
library(dplyr)

# Values pulled directly from beta_parts, not hardcoded
beta_data <- data.frame(
  Component = c("Similarity", "Turnover (βSIM)", "Nestedness (βSNE)"),
  Value = c(1 - beta_parts$beta.SOR, beta_parts$beta.SIM, beta_parts$beta.SNE)
)

beta_data$Component <- factor(beta_data$Component,
                               levels = c("Similarity", "Nestedness (βSNE)", "Turnover (βSIM)"))

beta_data <- beta_data %>%
  arrange(desc(Component)) %>%
  mutate(
    Percentage = (Value / sum(Value)) * 100,
    Label = paste0(round(Percentage, 1), "%"),
    ypos = cumsum(Value) - 0.5 * Value
  )

my_colors <- c(
  "Similarity"        = "#D2B48C",
  "Turnover (βSIM)"   = "#F67065",
  "Nestedness (βSNE)" = "#F3B0A1"
)

beta_pie <- ggplot(beta_data, aes(x = 2, y = Value, fill = Component)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y", start = 0) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = my_colors) +
  theme_void() +
  labs(
    title = "Beta diversity partitioning (Sørensen)",
    subtitle = "Relative contribution of turnover and nestedness",
    fill = "Component"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, color = "grey30"),
    legend.position = "right",
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA)
  ) +
  geom_text(aes(y = ypos, label = Label), color = "black", fontface = "bold", size = 4.5)

beta_pie

# ggsave(filename = "beta_diversity_partitioning.png", plot = beta_pie,
#        width = 5.5, height = 4, dpi = 300)
```

*##############################################################################*
