---
title: "Objective 3.3: Quantifying and disentangling the role of environment and space in shaping community structure"
output: html_notebook
---

Aim:
1. Redundancy Analysis (RDA)
2. Variance partitioning (environment, spatial structure, pure space)

Data required: 

site_species_monthly_quantitative_data_matrix.csv: 86 species remain after excluding Eriocaulon's species
site_environment_data_matrix.csv: site x month environmental data matrix (n = 32 rows = 8 sites x 4 months)
site_attributes_data_matrix.csv

*Aim 1: Redundancy Analysis (RDA)*

```{r}
library(dplyr)
library(vegan)
library(corrplot)
library(RColorBrewer)
library(sf)
library(adespatial)
```

1. Load and order the species and environmental data matrices:
```{r}
site_order  <- c("Rai", "Kas", "Raj", "Cha", "Udg", "Bor", "Kha", "Man")
month_order <- c("July", "Aug", "Sept", "Oct")

species_data <- read.csv("site_species_monthly_quantitative_data_matrix.csv", stringsAsFactors = FALSE) %>%
  mutate(Site = factor(Site, levels = site_order), Month = factor(Month, levels = month_order)) %>%
  arrange(Site, Month)

## site_environment_data_matrix.csv is transect-level (multiple rows per site x month);
## average across transects to get one row per site x month before use.
env_data_avg <- read.csv("site_environment_data_matrix.csv", stringsAsFactors = FALSE) %>%
  select(-any_of("Site_transect")) %>%
  mutate(Month = case_match(Month, "Jul" ~ "July", .default = Month)) %>%
  group_by(Site, Month) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  mutate(Site = factor(Site, levels = site_order), Month = factor(Month, levels = month_order)) %>%
  arrange(Site, Month)

stopifnot(all(species_data$Site == env_data_avg$Site), all(species_data$Month == env_data_avg$Month))
```

2. Subset numeric columns:
```{r}
species_data_final <- subset(species_data, select = -c(Site, Month))
env_data_final <- subset(env_data_avg, select = -c(Site, Month, Latitude, Longitude))
```

3. Hellinger-transform species abundance data and scale environmental data:
```{r}
## Note: Eriocaulon sp. are excluded from this abundance-based analysis, consistent
## with the main text. A supplementary version retaining Eriocaulon is run separately
## at the end of Objective 3.4.

species_hel <- decostand(species_data_final, method = "hellinger")
env_scaled <- as.data.frame(scale(env_data_final))
```

4. Correlation analysis to identify and remove collinear variables:
```{r}
cor_matrix <- cor(env_scaled, method = "pearson")
corrplot(cor_matrix, method = "number")  # Fig. S3
```

```{r}
# Based on the correlation matrix, wind speed, minimum temperature, and maximum temperature were each highly correlated (|r| > 0.8) with a retained variable and were excluded. 
# Retained: elevation, mean temperature (tavg), water vapour pressure (vapr), solar radiation (srad), and precipitation (prec).
env_sub <- subset(env_scaled, select = -c(tmax, tmin, wind))
```

5. Forward selection of significant environmental predictors:
```{r}
forward_model <- ordiR2step(
  rda(species_hel ~ 1, data = env_sub),
  scope = ~ Elevation + tavg + vapr + srad + prec,
  direction = "forward",
  permutations = 10000,
  R2scope = TRUE
)
forward_model$anova

selected_vars <- rownames(forward_model$anova)
selected_vars <- selected_vars[!grepl("All variables", selected_vars)]
selected_vars <- gsub("^\\+ ", "", selected_vars)
```

6. Run the final RDA using the forward-selected predictors:
```{r}
final_formula <- as.formula(paste("species_hel ~", paste(selected_vars, collapse = " + ")))
final_rda <- rda(final_formula, data = env_sub)

summary(final_rda)  # eigenvalues, species scores, site scores
```

7. Significance testing (constrained permutations):
```{r}
anova(final_rda, permutations = 9999)                # global model test
anova(final_rda, by = "axis", permutations = 9999)   # individual RDA axes
anova(final_rda, by = "terms", permutations = 9999)  # sequential (Type I): order-dependent
anova(final_rda, by = "margin", permutations = 9999) # marginal (Type III): unique contribution

# Repeated sampling of the same site across months violates independence, so permutations are additionally restricted within sites.
perm <- how(blocks = env_data_avg$Site)
anova(final_rda, by = "terms", permutations = perm)
```

8. Visualize the final model: sites with convex hulls and loading vectors:
```{r}
sites <- factor(env_data_avg$Site, levels = site_order)
site_levels <- levels(sites)
site_colors <- brewer.pal(n = length(site_levels), name = "Set2")

months <- factor(env_data_avg$Month, levels = month_order)
month_levels <- levels(months)
month_shapes <- c(21, 22, 24, 25)[months]

ordiplot(final_rda, type = "n", scaling = 2)

points(final_rda, display = "sites", pch = month_shapes, bg = site_colors[sites], col = "black")

poly_colors <- adjustcolor(site_colors, alpha.f = 0.5)
ordihull(final_rda, groups = sites, draw = "polygon",
         col = poly_colors, border = site_colors, lwd = 2, scaling = 2, label = FALSE)

env_arrows <- scores(final_rda, display = "bp", scaling = 2)
arrows(0, 0, env_arrows[, 1], env_arrows[, 2], length = 0.1, col = "darkblue", lwd = 1.5)
text(env_arrows[, 1], env_arrows[, 2], labels = rownames(env_arrows),
     col = "darkblue", cex = 0.9, pos = 2, font = 2)

legend("topright", legend = site_levels, fill = site_colors, border = "black",
       title = "Site", cex = 0.8, bty = "n")
legend("bottomright", legend = month_levels, pch = c(21, 22, 24, 25), pt.bg = "gray", col = "black",
       title = "Month", cex = 0.8, bty = "n")
```

*##############################################################################*

*Aim 2: Variance partitioning (environment, spatial structure, pure space)*

9. Project site coordinates to UTM (Zone 43N):
```{r}
site_attributes <- read.csv("site_attributes_data_matrix.csv", stringsAsFactors = FALSE) %>%
  mutate(Site = recode(Study.site,
    "Raireshwar" = "Rai", "Kaas" = "Kas", "Rajapuri" = "Raj", "Chalkewadi" = "Cha",
    "Udgiri" = "Udg", "Borbet" = "Bor", "Khamda dev" = "Kha", "Mangeli" = "Man"
  )) %>%
  select(Site, Latitude = Lat, Longitude = Long)
 
xy_coords <- env_data_avg %>%
  select(Site, Month) %>%
  left_join(site_attributes, by = "Site")
 
stopifnot(!any(is.na(xy_coords$Latitude)))  # confirm every site matched during recode
 
stopifnot(all(xy_coords$Longitude >= 72 & xy_coords$Longitude < 78))  # confirm Zone 43N
 
points_sf   <- st_as_sf(xy_coords, coords = c("Longitude", "Latitude"), crs = 4326)
proj_string <- "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs"
points_utm  <- st_transform(points_sf, crs = proj_string)
 
XY <- as.data.frame(st_coordinates(points_utm)) / 1000  # rescale to km
names(XY) <- c("X", "Y")
```

10. Distance-based Moran's Eigenvector Maps (dbMEMs):
```{r}
dbmem_vars <- dbmem(XY, silent = FALSE)
 
rda_mems <- rda(species_hel ~ ., data = dbmem_vars)
summary(rda_mems)
 
forward_mems <- ordiR2step(
  rda(species_hel ~ 1, data = dbmem_vars),
  scope = formula(rda_mems),
  direction = "forward",
  permutations = 10000
)
forward_mems$anova
 
mems_selected <- rownames(forward_mems$anova)
mems_selected <- mems_selected[!grepl("All variables", mems_selected)]
mems_selected <- gsub("^\\+ ", "", mems_selected)
mems_vars <- dbmem_vars[, mems_selected, drop = FALSE]
```

11. Perform variance partitioning: environment, spatial structure, space:
```{r}
env_vars <- env_sub[, selected_vars]
 
varpart_result <- varpart(species_hel, env_vars, mems_vars, XY)
varpart_result
 
plot(varpart_result,
     Xnames = c("Env.", "MEM", "Space"),
     bg = c("#F67065", "#335C8A", "#6AA6B4"),
     col = "black", lwd = 1.2, id.size = 0, cex = 1)
title("Variance partitioning of species assemblages", font.main = 2, cex.main = 1.2)
```

12. Test the significance of each component:
```{r}
anova.cca(rda(species_hel, env_vars))
RsquareAdj(rda(species_hel, env_vars))$adj.r.squared
 
anova.cca(rda(species_hel, mems_vars))
RsquareAdj(rda(species_hel, mems_vars))$adj.r.squared
 
anova.cca(rda(species_hel, XY))
RsquareAdj(rda(species_hel, XY))$adj.r.squared
```

*##############################################################################*
