---
title: "Objective 2.2: Testing distance-decay relationship (Mantel test)"
output: html_notebook
---

Aim:
1. To test whether compositional dissimilarity between sites increases with geographic distance, using Mantel tests on both Jaccard (presence/absence) and Bray-Curtis (abundance) dissimilarity matrices.
2. To test the distance-dacy relationship at genus and family level (for taxonomic resolution sensitivity check)

Data required: 

geographic_distance_matrix.csv, 
jaccard_dissimilarity_matrix.csv,
bray_curtis_dissimilarity_matrix.csv
site_genus_data_matrix.csv,
site_family_data_matrix.csv

*Aim 1: To test whether compositional dissimilarity between sites increases with geographic distance, using Mantel tests on both Jaccard (presence/absence) and Bray-Curtis (abundance) dissimilarity matrices.*

--- Species-level: Jaccard dissimilarity vs. geographic distance ---

1. Load data:
```{r}
distance_data <- read.csv("geographic_distance_matrix.csv")
jaccard_data  <- read.csv("jaccard_dissimilarity_matrix.csv")

distance_data <- na.omit(distance_data)
```

2. Convert to distance objects:
```{r}
dist_matrix <- as.matrix(distance_data[, -1])
jac_matrix  <- as.matrix(jaccard_data[, -1])
jac_matrix  <- 1 - jac_matrix  # convert similarity to dissimilarity, if stored as similarity

dist_matrix <- as.dist(dist_matrix)
jac_matrix  <- as.dist(jac_matrix)
```

3. Run Mantel test:
```{r}
library(vegan)

mantel_jaccard <- mantel(jac_matrix, dist_matrix, method = "pearson", permutations = 9999)
mantel_jaccard
```

4. Scatter plot:
```{r}
library(ggplot2)

dist_values <- as.numeric(as.matrix(dist_matrix)[lower.tri(as.matrix(dist_matrix))])
jac_values  <- as.numeric(as.matrix(jac_matrix)[lower.tri(as.matrix(jac_matrix))])

data_mantel_jaccard <- data.frame(distance = dist_values, dissimilarity = jac_values)

label_jaccard <- paste0("Mantel's r = ", round(mantel_jaccard$statistic, 3),
                         ", P = ", round(mantel_jaccard$signif, 3))

jaccard_plot <- ggplot(data_mantel_jaccard, aes(x = distance, y = dissimilarity)) +
  geom_smooth(method = "lm", col = "black", linewidth = 1.5) +
  geom_point(color = "#F67065", size = 3) +
  ggtitle("Community dissimilarity vs. geographical distance: Mantel test") +
  xlab("Geographical distance (km)") +
  ylab("Jaccard dissimilarity index") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.background = element_rect(fill = "transparent", color = NA)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1.5,
           label = label_jaccard, size = 4.5, fontface = "italic", color = "black")

jaccard_plot

# ggsave(filename = "similarity_vs_distance_jaccard.png", plot = jaccard_plot,
#        width = 6, height = 4, dpi = 300)
```

--- Species-level: Bray-Curtis dissimilarity vs. geographic distance ---

5. Load data:
```{r}
bray_data <- read.csv("bray_curtis_dissimilarity_matrix.csv")
```

6. Convert to distance objects:
```{r}
bray_matrix <- as.matrix(bray_data[, -1])
bray_matrix <- 1 - bray_matrix  # convert similarity to dissimilarity, if stored as similarity
bray_matrix <- as.dist(bray_matrix)
```

7. Run Mantel test:
```{r}
mantel_bray <- mantel(bray_matrix, dist_matrix, method = "pearson", permutations = 9999)
mantel_bray
```

8. Scatter plot:
```{r}
bray_values <- as.numeric(as.matrix(bray_matrix)[lower.tri(as.matrix(bray_matrix))])

data_mantel_bray <- data.frame(distance = dist_values, dissimilarity = bray_values)

label_bray <- paste0("Mantel's r = ", round(mantel_bray$statistic, 3),
                      ", P = ", round(mantel_bray$signif, 3))

bray_plot <- ggplot(data_mantel_bray, aes(x = distance, y = dissimilarity)) +
  geom_smooth(method = "lm", col = "black", linewidth = 1.5) +
  geom_point(color = "#F67065", size = 3) +
  ggtitle("Community dissimilarity vs. geographical distance: Mantel test") +
  xlab("Geographical distance (km)") +
  ylab("Bray-Curtis dissimilarity index") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.background = element_rect(fill = "transparent", color = NA)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1.5,
           label = label_bray, size = 4.5, fontface = "italic", color = "black")

bray_plot

# ggsave(filename = "similarity_vs_distance_bray.png", plot = bray_plot,
#        width = 6, height = 4, dpi = 300)
```

--- Supplementary figure: Hierarchical clustering and similarity heatmaps ---

9. Load dissimilarity matrices and convert to similarity for display:
```{r}
library(pheatmap)
library(viridis)
library(dendextend)

jaccard_diss <- read.csv("jaccard_dissimilarity_matrix.csv", row.names = 1)
bray_diss    <- read.csv("bray_curtis_dissimilarity_matrix.csv", row.names = 1)

# Similarity is more intuitive for visual interpretation than dissimilarity:
# higher values = more similar = darker cells
jaccard_sim <- 1 - as.matrix(jaccard_diss)
bray_sim    <- 1 - as.matrix(bray_diss)
```

10. Hierarchical clustering (average linkage) and manual reordering:
```{r}
jacc_distance <- as.dist(jaccard_diss)
bray_distance <- as.dist(bray_diss)

jacc_hc <- hclust(jacc_distance, method = "average")
bray_hc <- hclust(bray_distance, method = "average")

jacc_dendrogram <- as.dendrogram(jacc_hc)
bray_dendrogram <- as.dendrogram(bray_hc)

# Reorder dendrogram branches to reflect the north-to-south site sequence
desired_order <- c("Man", "Kha", "Bor", "Udg", "Cha", "Raj", "Kas", "Rai")

jacc_dendrogram_ordered <- rotate(jacc_dendrogram, order = desired_order)
bray_dendrogram_ordered <- rotate(bray_dendrogram, order = desired_order)

jacc_hc_rotated <- as.hclust(jacc_dendrogram_ordered)
bray_hc_rotated <- as.hclust(bray_dendrogram_ordered)
```

11. Panel A: Jaccard similarity heatmap:
```{r}
jaccard_pheatmap <- pheatmap(jaccard_sim,
                    cluster_rows = jacc_hc_rotated,
                    cluster_cols = jacc_hc_rotated,   # cluster both axes so the diagonal stays interpretable
                    display_numbers = TRUE,
                    number_color = "black",
                    fontsize_number = 12,
                    color = viridis::viridis(100, direction = -1),
                    border_color = "white",
                    main = "Jaccard similarity index between study sites",
                    fontsize_row = 10,
                    fontsize_col = 10,
                    number_format = "%.2f",
                    treeheight_row = 80,
                    treeheight_col = 0,   # hide column dendrogram, keep only the row dendrogram
                    angle_col = 45,
                    show_rownames = TRUE,
                    show_colnames = TRUE,
                    legend = TRUE)

# pheatmap(jaccard_sim, ..., filename = "fig_S8A_jaccard_heatmap.png",
#          width = 7, height = 5)
```

12. Panel B: Bray-Curtis similarity heatmap:
```{r}
bray_pheatmap <- pheatmap(bray_sim,
         cluster_rows = bray_hc_rotated,
         cluster_cols = bray_hc_rotated,
         display_numbers = TRUE,
         number_color = "black",
         fontsize_number = 12,
         color = viridis::viridis(100, direction = -1),
         border_color = "white",
         main = "Bray-Curtis similarity index between study sites",
         fontsize_row = 10,
         fontsize_col = 10,
         number_format = "%.2f",
         treeheight_row = 80,
         treeheight_col = 0,
         angle_col = 45,
         show_rownames = TRUE,
         show_colnames = TRUE,
         legend = TRUE)

# pheatmap(bray_sim, ..., filename = "fig_S8B_bray_heatmap.png",
#          width = 7, height = 5)
```

*##############################################################################*

*Aim 2: To test the distance-dacy relationship at genus and family level (for taxonomic resolution sensitivity check)*

--- Genus-level: Bray-Curtis dissimilarity vs. geographic distance ---

9. Load data:
```{r}
genus_final <- read.csv("site_genus_data_matrix.csv", row.names = 1)
```

10. Hellinger transformation and Bray-Curtis dissimilarity:
```{r}
genus_hell <- decostand(genus_final, method = "hellinger")

genus_diss <- vegdist(genus_hell, method = "bray", binary = FALSE)
genus_diss <- as.matrix(genus_diss)
```

11. Mantel test:
```{r}
genus_diss_d <- as.dist(genus_diss)

mantel_genus <- mantel(genus_diss_d, dist_matrix, method = "pearson", permutations = 9999)
mantel_genus
```

12. Scatter plot:
```{r}
genus_diss_vec <- as.vector(genus_diss_d)

genus_mantel <- data.frame(
  Geographic_distance = dist_values,
  Genus_dissimilarity = genus_diss_vec
)

label_genus <- paste0("Mantel's r = ", round(mantel_genus$statistic, 3),
                       ", P = ", round(mantel_genus$signif, 3))

genus_plot <- ggplot(genus_mantel, aes(x = Geographic_distance, y = Genus_dissimilarity)) +
  geom_smooth(method = "lm", col = "black", linewidth = 1.5) +
  geom_point(color = "#F67065", size = 3) +
  ggtitle("Community dissimilarity vs. geographical distance: Genus") +
  xlab("Geographical distance (km)") +
  ylab("Bray-Curtis dissimilarity index: Genus-level") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.background = element_rect(fill = "transparent", color = NA)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1.5,
           label = label_genus, size = 4.5, fontface = "italic", color = "black")

genus_plot

# ggsave(filename = "similarity_vs_distance_genus.png", plot = genus_plot,
#        width = 6, height = 4, dpi = 300)
```

--- Family-level: Bray-Curtis dissimilarity vs. geographic distance ---

13. Load data:
```{r}
family_final <- read.csv("site_family_data_matrix.csv", row.names = 1)
```

14. Hellinger transformation and Bray-Curtis dissimilarity:
```{r}
family_hell <- decostand(family_final, method = "hellinger")

family_diss <- vegdist(family_hell, method = "bray", binary = FALSE)
family_diss <- as.matrix(family_diss)
```

15. Mantel test:
```{r}
family_diss_d <- as.dist(family_diss)

mantel_family <- mantel(family_diss_d, dist_matrix, method = "pearson", permutations = 9999)
mantel_family
```

16. Scatter plot:
```{r}
family_diss_vec <- as.vector(family_diss_d)

family_mantel <- data.frame(
  Geographic_distance = dist_values,
  Family_dissimilarity = family_diss_vec
)

label_family <- paste0("Mantel's r = ", round(mantel_family$statistic, 3),
                        ", P = ", round(mantel_family$signif, 3))

family_plot <- ggplot(family_mantel, aes(x = Geographic_distance, y = Family_dissimilarity)) +
  geom_smooth(method = "lm", col = "black", linewidth = 1.5) +
  geom_point(color = "#F67065", size = 3) +
  ggtitle("Community dissimilarity vs. geographical distance: Family") +
  xlab("Geographical distance (km)") +
  ylab("Bray-Curtis dissimilarity index: Family-level") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.background = element_rect(fill = "transparent", color = NA)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1.5,
           label = label_family, size = 4.5, fontface = "italic", color = "black")

family_plot

# ggsave(filename = "similarity_vs_distance_family.png", plot = family_plot,
#        width = 6, height = 4, dpi = 300)
```

*##############################################################################*
