---
title: "Objective 3.1: Characterizing environmental gradients across plateaus (PCA)"
output: html_notebook
---

*Aim: To perform PCA analysis using monthly environmental data across sites*

Data required: site_environment_data_matrix.csv: site x month environmental data matrix (n = 32 rows = 8 sites x 4 months)

```{r}
library(dplyr)
library(corrplot)
library(ggplot2)
```

1. Load and average the site x month environmental data matrix:
```{r}
data <- read.csv("site_environment_data_matrix.csv", stringsAsFactors = FALSE)

avg_data <- data %>%
  group_by(Site, Month) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)), .groups = "drop") %>%
  arrange(factor(Site, levels = unique(data$Site)))

# Subset numeric environmental variables and standardize
data_final <- subset(avg_data, select = -c(Site, Month, Latitude, Longitude))
data_scaled <- as.data.frame(scale(data_final))
```

2. Correlation analysis to identify and remove collinear variables:
```{r}
cor_matrix <- cor(data_scaled, method = "pearson")

new_labels <- c("Elevation", "Max. temp.", "Min. temp.", "Avg. temp.",
                 "Water vapour", "Wind speed", "Solar rad.", "Precipitation")
rownames(cor_matrix) <- new_labels
colnames(cor_matrix) <- new_labels

corrplot(cor_matrix, method = "number", tl.col = "black", tl.srt = 45)  # Fig. S3
```

Based on the correlation matrix, wind speed, minimum temperature, and maximum temperature were each highly correlated (|r| > 0.8) and were excluded. 
Retained: elevation, mean temperature (tavg), water vapour pressure (vapr), solar radiation (srad), and precipitation (prec).

3. Perform PCA on the reduced set of environmental variables:
```{r}
data_pca <- subset(data_scaled, select = -c(tmax, tmin, wind))

pca_result <- prcomp(data_pca, center = FALSE, scale. = FALSE)  # already centered/scaled above

pca_summary <- summary(pca_result)
var_explained <- pca_summary$importance["Proportion of Variance", ] * 100  # dynamic, not hardcoded

cat(sprintf(
  "PC1 explains %.2f%% and PC2 explains %.2f%% of total variation (first two axes: %.2f%%)\n",
  var_explained["PC1"], var_explained["PC2"],
  var_explained["PC1"] + var_explained["PC2"]
))

pca_result$rotation  # Table S5: loadings across all axes

pca_scores <- as.data.frame(pca_result$x)
pca_scores$Site  <- factor(avg_data$Site, levels = unique(avg_data$Site))
pca_scores$Month <- factor(avg_data$Month, levels = c("Jul", "Aug", "Sept", "Oct"))
```


4. Extract loadings for PC1/PC2, scaled for display:
```{r}
loadings <- as.data.frame(pca_result$rotation[, 1:2])
loadings$Variable <- rownames(loadings)

arrow_scale <- 4
loadings$PC1 <- loadings$PC1 * arrow_scale
loadings$PC2 <- loadings$PC2 * arrow_scale

## 5. Visualize the PCA ordination with loading vectors (Fig. 3B) ------------

site_colors <- c(
  "Rai" = "#223A70", "Kas" = "#335C8A", "Raj" = "#4B7CA3", "Cha" = "#6AA6B4",
  "Udg" = "#5FA782", "Bor" = "#88B04B", "Kha" = "#E0C54F", "Man" = "#D4A937"
)
site_labels <- c(
  "Rai" = "Raireshwar", "Kas" = "Kaas", "Raj" = "Rajapuri", "Cha" = "Chalkewadi",
  "Udg" = "Udgiri", "Bor" = "Borbet", "Kha" = "Khamda dev", "Man" = "Mangeli"
)

pc1_lab <- sprintf("PC1 (Explained %.2f%% variation)", var_explained["PC1"])
pc2_lab <- sprintf("PC2 (Explained %.2f%% variation)", var_explained["PC2"])

pca_plot_loadings <- ggplot(pca_scores, aes(x = PC1, y = PC2, label = Site)) +
  geom_polygon(aes(group = Site, fill = Site, color = Site), linewidth = 0.8, alpha = 0.8) +
  geom_point(aes(fill = Site, shape = Month), color = "black", size = 3, show.legend = c(fill = FALSE)) +
  scale_shape_manual(values = c("Jul" = 21, "Aug" = 22, "Sept" = 24, "Oct" = 25)) +
  scale_fill_manual(values = site_colors, labels = site_labels) +
  scale_color_manual(values = site_colors, labels = site_labels) +
  geom_segment(data = loadings, inherit.aes = FALSE,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.25, "cm")), color = "black", linewidth = 0.8) +
  geom_text(data = loadings, inherit.aes = FALSE,
            aes(x = PC1, y = PC2, label = Variable),
            color = "black", size = 4, nudge_x = 0.37, nudge_y = 0.15) +
  guides(
    fill = guide_legend(order = 1, title = "Site"),
    color = "none",
    shape = guide_legend(order = 2, title = "Month")
  ) +
  theme_bw() +
  labs(title = "PCA of environmental conditions for months across sites",
       x = pc1_lab, y = pc2_lab) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        axis.title.x = element_text(face = "bold", size = 14),
        axis.title.y = element_text(face = "bold", size = 14),
        legend.title = element_text(size = 13, face = "bold"),
        legend.text  = element_text(size = 12))

pca_plot_loadings
```

*##############################################################################*
