---
title: "Objective 1.2: Testing species-area relationship (Spearman correlation)"
output: html_notebook
---

*Aim: To test whether species richness increases with plateau area, using Spearman's rank correlation.*

This method assesses monotonic relationships without assuming linearity, and is robust to outliers, making it more
appropriate here than Pearson's correlation given the small sample size (n = 8) and the wide range in plateau area.

Data required: site_attributes_matrix.csv

1. Load data:
```{r}
library(dplyr)

site_attributes <- read.csv("site_attributes_data_matrix.csv")

rich_area <- site_attributes %>%
  select(Study.site, Species.richness, Area) # Area is in sq. km.

print(rich_area)
```

2. Run Spearman's rank correlation (all eight sites included):
```{r}
spearman_result <- cor.test(rich_area$Species.richness, rich_area$Area,
                             method = "spearman")

spearman_result
```

Result:

Correlation coefficient (rho): 0.323 (weak positive correlation)
p-value: 0.434 (not statistically significant)

3. Scatter plot of species richness vs. area:
```{r}
library(ggplot2)
library(ggrepel)

# Build the annotation label directly from the test result, so the plot
# can never fall out of sync with the reported statistics
spearman_label <- paste0(
  "rho = ", round(spearman_result$estimate, 3),
  ", P = ", round(spearman_result$p.value, 3)
)

spearman_plot <- ggplot(rich_area, aes(x = Area, y = Species.richness)) +

  geom_smooth(method = "lm", color = "black", linewidth = 1.5, se = TRUE) +

  geom_point(color = "#F67065", size = 3) +

  geom_text_repel(aes(label = Study.site), size = 3,
                   box.padding = 0.3,
                   point.padding = 0.3) +

  theme_bw() +

  theme(plot.background = element_rect(fill = "transparent", color = NA),
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold", size = 12)) +

  ggtitle("Spearman correlation between species richness and area") +
  xlab("Area (sq. km)") + ylab("Species richness") +

  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1.5,
           label = spearman_label,
           size = 4.5, fontface = "italic", color = "black")

spearman_plot

# Save the plot
# ggsave(filename = "species_richness_vs_area.png", plot = spearman_plot,
#        width = 6, height = 4, dpi = 300)
```

*##############################################################################*
